# Log-Gaussian Cox process {#lgcp}

A Log-Gaussian Cox Process (LGCP) is a doubly stochastic spatial point process. In the simplest case, the intensity of the point process over space is given by:
$$\Lambda(s) = \text{exp}(\beta_0 + G(s) + \epsilon)$$
where $\beta_0$ is a constant, known as the intercept, $G(s)$ is a Gaussian Markov Random Field (GMRF) and $\epsilon$ an error term. 

It is conventional to use Matérn covariance function to define the covariance of the random field. This takes two parameters $\tau$ and $\kappa$, commonly reported as $r=\frac{\sqrt{8}}{\kappa}$ and $\sigma=\frac{1}{\sqrt{4\pi\kappa^2\tau^2}}$, where $r$ is the range and $\sigma$ is the standard deviation. 

## The `fit_lgcp()` function

```{r}
args(fit_lgcp)
```


### Fitting a spatial only LGCP

```{r}
data(xyt, package = "stelfi")
domain <- sf::st_as_sf(xyt$window)
locs <- data.frame(x = xyt$x, y = xyt$y)
bnd <- INLA::inla.mesh.segment(as.matrix(sf::st_coordinates(domain)[, 1:2]))
smesh <- INLA::inla.mesh.2d(boundary = bnd,
                                max.edge = 0.75, cutoff = 0.3)
fit <- fit_lgcp(locs = locs, sf = domain, smesh = smesh,
                                parameters = c(beta = 0, log_tau = log(1),
                                               log_kappa = log(1)))
get_coefs(fit)
```


```{r}
get_fields(fit, smesh) |>
    show_field(smesh = smesh, sf = domain) + ggplot2::theme_void()
show_lambda(fit, smesh = smesh, sf = domain) + ggplot2::theme_void()
```




## The `sim_lgcp()` function

**Simulating a spatiotemporal LGCP**

**Option 1** using `sim_lgcp`

```{r}
parameters <- c(beta = 1, log_tau = log(1), log_kappa = log(1), atanh_rho = 0.2)
simdata <- sim_lgcp(parameters = parameters, sf = domain, smesh = smesh, tmesh = tmesh)
simdata$x[,1] |>
    show_field(smesh = smesh) +
    ggplot2::theme_void()
```

**Option 2** directly from the fitted model

```{r}
simdata <- fit$simulate()
simdata$x[,1] |>
    show_field(smesh = smesh) +
    ggplot2::theme_void()
```

