---
title: "Modelling retweets with a Hawkes process"
author: "Charlotte M. Jones-Todd"
date: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling retweets with a Hawkes process}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE,message=FALSE,results='asis'}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,cache = TRUE)
```


## Fitting a Hawkes process to Twitter data

A [NIWA](https://niwa.co.nz/) scientist [found a working USB in the scat of a leopard seal](https://www.nzherald.co.nz/nz/news/article.cfm?c_id=1&objectid=12201147), they then [tweeted about it](https://twitter.com/niwa_nz/status/1092610541401587712) in the hopes of finding its owner.

Below we use the `stelfi` package to model the times of retweets as a self-exciting (Hawkes) process.

```{r load data,results='hide',message=FALSE}
library(stelfi)
data(retweets_niwa)
head(retweets_niwa)
```

```{r sort data}
## numeric time stamps
times <- sort(as.numeric(difftime(retweets_niwa ,min(retweets_niwa),units = "mins")))
```

```{r plot hist, echo = TRUE, fig.cap="Observed counts of retweet times.",fig.height=5,fig.width=8}
# NIWA seal first tweet according to Twitter
start <- lubridate::ymd_hms("2019-02-05 02:27:07")
## NIWA seal found owner tweet according to Twitter
end <- lubridate::ymd_hms("2019-02-07 06:50:08")
## hist
hist(retweets_niwa, breaks = "hours", axes = FALSE, 
     xlab = "", ylab = "",main = "", col = "grey",border = "grey",freq = TRUE)
axis(2,at = c(0,250))
mtext(2, line = 0, text = "Number of retweets per hour",cex = 0.8)
mtext(1,line = 1, at = c(start,end),text = c("NIWA \n tweeted","USB owner \n found"),cex = 0.9)
```

### Model fitting using `TMB`

Before using the  `TMB` templates in `stelfi` you should use `compile.stelfi()` to compile them.

```{r compile,results = "hide"}
compile.stelfi()
```

```{r fit model,results='hide',cache=TRUE}
params <- c(mu = 9,alpha = 3,beta = 10)
## must have compiled TMB templates first use compile.stelfi()
fit <- fit_hawkes(times = times,parameters = params) 
```


```{r params}
## print out estimated parameters
fit
par <- fit$par.fixed
par
```



```{r plot, echo = FALSE,fig.height=7,fig.width=9,fig.cap="Fitted intensity (top plot) and observed counts of retweets (bottom)."}
p <- seq(min(times),max(times),length.out = 1000)
intenst <- hawke_intensity(mu = par[1],alpha = par[2],beta = par[3],times = times, p = p)
layout(matrix(c(1,1,1,1,1,1,2,2,2),byrow = TRUE, ncol = 3))
par(mar = c(1,5,1,2))
plot(times,rep(0,length(retweets_niwa)),pch = "|", yaxt = "n",ylab = "", xlab = "", ylim = c(0,max(intenst)),
     xaxt = "n",cex = 0.6)
mtext(2,line = 0,text = "Rate of retweets per minute", cex = 0.9)
lines(p,intenst,col = "grey")
axis(2,at = round(c(0,max(intenst)),2),las = 2)
const <- length(times)/max(times)
abline(h = const, lty = 2)
text(1000,const + 0.5, paste("Constant rate of",round(const,2),"retweets per minute"))
arrows(1000,const + 0.45, 1000,const,length = 0.05)
day2 <- start + 24*60*60
day3 <- day2 + 24*60*60
abline(v = as.numeric(difftime(day2,min(retweets_niwa),unit = "mins")))
text(as.numeric(difftime(day2,min(retweets_niwa),unit = "mins")) - 100,4,"24hrs since tweet",srt = 90)
abline(v = as.numeric(difftime(day3,min(retweets_niwa),unit = "mins")))
text(as.numeric(difftime(day3,min(retweets_niwa),unit = "mins")) - 100,4,"48hrs since tweet",srt = 90)
par(mar = c(2,5,0,2))
## hist
hist(retweets_niwa, breaks = "mins", axes = FALSE, 
     xlab = "", ylab = "",main = "", col = "grey",border = "grey",freq = TRUE)
axis(2,at = c(0,11))
mtext(2, line = 0, text = "Number of retweets per minute",cex = 0.8)
mtext(1,line = 1, at = c(start,end),text = c("NIWA \n tweeted","USB owner \n found"),cex = 0.9)
```
