---
title: ""
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE,message=FALSE,results='asis'}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,cache = TRUE, message = FALSE,warning=FALSE)
library(kableExtra)
```
To install the `R` package `stelfi` run `devtools::install_github("cmjt/stelfi")`.

```{r lib}
library(stelfi)
```

# NZ murders

## The Data

```{r data}
data(murders_nz)
dim(murders_nz)
head(murders_nz)
```


```{r eda, echo = FALSE,results='asis'}
kable(table(murders_nz$Region), format = "pipe", col.names = c("Region","Number"),caption = "Number of murders by province") 
```
*Number of murders by province.*


```{r rate}
data(nz) ## SpatialPolygonsDataFrame of NZ (NZTM projection)
area_nz <- sum(raster::area(nz)) ## (m)
area_nzkm2 <- area_nz/1000^2 ## according to Google NZ is 268,021 km2
area_nzkm2
spatial_murder_rate <- nrow(murders_nz)/area_nzkm2 
temporal_murder_rate <- nrow(murders_nz)/length(table(murders_nz$Year)) 
st_murder_rate <- (nrow(murders_nz)/area_nzkm2)/length(table(murders_nz$Year)) 
```

There are on average `r round(temporal_murder_rate,1)` murders per year. The rate of murders per $\text{km}^2$ across NZ is roughly `r round(spatial_murder_rate,3)`. 

### Subset data to Waikato & transform to NZTM

Transform `data.frame` to `SpatialPointsDataFrame` and project Longitude and Latitude to NZTM.

```{r sp}
murders_sp <- subset(murders_nz, murders_nz$Region == "Waikato")
## project longitude & latitude to NZTMs
coordinates(murders_sp) <- c("Longitude","Latitude")
proj4string(murders_sp) <- CRS("+proj=longlat +datum=WGS84")
murders_sp <-  spTransform(murders_sp, 
                           CRS("+proj=nzmg +lat_0=-41.0 +lon_0=173.0 +x_0=2510000.0 +y_0=6023150.0 +ellps=intl +units=m"))
waikato <- nz[nz$NAME_1 == "Waikato",]
```

```{r edatable, echo = FALSE,results='asis'}
kable(table(murders_sp$Cause_cat), format = "pipe", col.names = c("Region","Number"),caption = "Number of murders by category in Waikato 2004--2019") 
```
*Number of murders by category in Waikato 2004--2019.*

```{r plot,results='asis',fig.width=10,fig.height=10,echo = FALSE,warning=FALSE,message=FALSE}
plot(waikato)
plot(murders_sp, add = TRUE)
```
*Locations of recorded (n = 108) murders in Waikato 2004--2019*

## log-Gaussian Cox process

### `stelfi` as a wrapper for `INLA`

#### Creating the mesh

Typically when analysing point pattern data the point locations are not specified as the mesh nodes (i.e., locations are not given as an argument to `inla.mesh.2d()`). Instad we can supply the coordinates of the point pattern window (domain).

```{r create mesh}
## mesh max.edge on the same scale as the coords (NZTMs)
mesh <- inla.mesh.2d(loc.domain = broom::tidy(waikato)[,1:2],
                     max.edge = c(11000,20000), cutoff = 15000)
## areas/weights at each mesh nodes
weights <- stelfi::get_weights(mesh = mesh, sp = waikato, plot = FALSE)
```

```{r using stelfi}
## Spatial only
fit <- fit_lgcp_inla(mesh = mesh, locs = coordinates(murders_sp), sp = waikato)
summary(fit)
## fixed effects
fit$summary.fixed
## expected number of murders at each mesh node
ins <- which(weights != 0)
en <- exp(as.numeric(fit$summary.fixed[1]))*weights[ins]
sum(en) ## expected number across Waikato, observed 108
```

```{r random fields, fig.width=15,fig.height=15}
fields <- stelfi::get_fields(fit, mesh, mean = TRUE)
grfs <- fields[[1]]
show_field(grfs, mesh, dims = c(300,300),
		 col = RColorBrewer::brewer.pal(9, "Blues"), sp = waikato,
           	 rast = FALSE, legend = TRUE, legend.only = FALSE)

```
*Estimated mean of the assumed Gaussian Markov Random Field* 


```{r random fields sd, fig.width=15,fig.height=15}
fields <- stelfi::get_fields(fit, mesh, mean = FALSE)
grfsd <- fields[[1]]
show_field(grfsd, mesh, dims = c(300,300),
		 col = RColorBrewer::brewer.pal(9, "Blues"), sp = waikato,
           	 rast = FALSE, legend = TRUE, legend.only = FALSE)

```
*Standard error of the assumed Gaussian Markov Random Field* 


### "Raw" `INLA`

**Steps below closely follow this [INLA-SPDE tutorial](https://becarioprecario.bitbucket.io/spde-gitbook/ch-lcox.html)**.

The SPDE approach for point pattern analysis defines the model at the nodes of the mesh. To fit the log-Cox point process model these points are considered as integration points. The method in Simpson et al. (2016) defines the expected number of events to be proportional to the area around the node (the areas of the polygons in the dual mesh, see below). This means that at the nodes of the mesh with larger triangles, there are also larger expected values.




```{r dual mesh,cache = TRUE,message = FALSE, fig.width=10,fig.height=10}
## these weights are the areas of each mesh triangle,
## required for the "exposure" aspect of the LGCP.
## the weights are set to zero if outside the study region
## as here the they should have no contribution. 
## the sum of the weights is the area of the study region
weights <- stelfi::get_weights(mesh = mesh, sp = waikato, plot = TRUE)
```
*Delauney triangulation of the domain (white) overlain on the Voronoi diagram representing the weights (area surrounding) of each mesh node (diamonds). Observations are plotted as circles, mesh nodes outwith the domain are shown in white.*


```{r plot weights,echo = FALSE, fig.width=10,fig.height=10}
## function below is a wrapper for
## the book.dual.mesh() function supplied
## in the INLA-SPDE tutorial mentioned above
dual_mesh <- stelfi:::inla.mesh.dual(mesh)
## expected number per mesh node
ins <- which(weights != 0)
pal <- scales::col_numeric("Blues",weights[ins]/1000^2)
cols <- pal(weights[ins]/1000^2)
plot(dual_mesh[ins], col = cols)
plot(waikato, add = TRUE)
fields::image.plot(zlim = range(weights[ins]/1000^2),col = sort(cols, TRUE), legend.only = TRUE, legend.width = 2, nlevel = 10, legend.shrink = 0.5)
```
*Voronoi diagram of the weights (shown as areas in km2 around each mesh node).*

```{r model fitting}
## number of mesh nodes
nodes <- mesh$n
## define model
spde <- inla.spde2.pcmatern(mesh = mesh,
  # PC-prior on range: P(practic.range < 5) = 0.9
  prior.range = c(5, 0.9),
  # PC-prior on sigma: P(sigma > 1) = 0.005
  prior.sigma = c(1, 0.005))
## vector for observations
y.pp <- rep(0:1, c(nodes, nrow(murders_sp)))
## exposure (E)
e.pp <- c(weights, rep(0, nrow(murders_sp)))
## integration points
imat <- Diagonal(nodes, rep(1, nodes))
## projection matrix for observed points
lmat <- inla.spde.make.A(mesh, coordinates(murders_sp))
## entire projection matrix
A.pp <- rbind(imat, lmat)
## data stack
stk.pp <- inla.stack(
  data = list(y = y.pp, e = e.pp), 
  A = list(1, A.pp),
  effects = list(list(b0 = rep(1, nodes + nrow(murders_sp))), 
                 list(i = 1:nodes)),
  tag = 'pp')
## fit model
pp.res <- inla(y ~ 0 + b0 + f(i, model = spde), 
  family = 'poisson', data = inla.stack.data(stk.pp), 
  control.predictor = list(A = inla.stack.A(stk.pp)), 
  E = inla.stack.data(stk.pp)$e)
summary(pp.res)
## fixed effects
pp.res$summary.fixed
## expected number of murders at each mesh node
ins <- which(weights != 0)
en <- exp(as.numeric(pp.res$summary.fixed[1]))*weights[ins]
sum(en) ## expected number across Waikato, observed 108
```


```{r resp,echo = FALSE, fig.width=15,fig.height=15}
## projected
fields <- stelfi::get_fields(pp.res, mesh, mean = TRUE)
grfs <- fields[[1]]
show_field(grfs, mesh, dims = c(300,300),
		 col = RColorBrewer::brewer.pal(9, "Blues"), sp = waikato,
           	 rast = FALSE, legend = TRUE, FALSE)
```
*Estimated mean of the assumed Gaussian Markov Random Field*

##### Adding a covariate

```{r, message = FALSE}
## include covariates
## The covariate shapefile used can be downloaded from
## https://koordinates.com/from/datafinder.stats.govt.nz/layer/8437/data/
## the code below assumes a single .shp (above) file is
## in a directory data/ relative to your working directory
file <- list.files("data",pattern = ".shp", full = TRUE)
layer <- rgdal::ogrListLayers(file)
pop <- rgdal::readOGR(file, layer = layer)
pop <- spTransform(pop, CRS("+proj=nzmg +lat_0=-41.0 +lon_0=173.0 +x_0=2510000.0 +y_0=6023150.0 +ellps=intl +units=m"))
pop_mesh <- sp::over(SpatialPoints(mesh$loc[,1:2], proj4string = CRS(proj4string(murders_sp))),pop)$Population
## will obviously be NA at mesh nodes outside NZ
```

```{r inference,echo = FALSE, fig.width=15,fig.height=15}
pal <- scales::col_numeric("Blues", pop_mesh[ins])
cols <- pal(pop_mesh[ins])
plot(dual_mesh[ins], col = cols)
plot(waikato, add = TRUE)
fields::image.plot(zlim = range(pop_mesh[ins], na.rm = TRUE),
                   col = sort(cols, TRUE), legend.only = TRUE, 
                   legend.width = 2, nlevel = 10, legend.shrink = 0.5)
```
*Voronoi diagram of the covariate (population) per mesh node.*


```{r fit cov model}
pop_obs <- sp::over(murders_sp,pop)$Population
## population density covariate c at mesh nodes and then obs locations 
covs <- data.frame(pop = c(pop_mesh, pop_obs))
## data stack
stk.cov <- inla.stack(
  data = list(y = y.pp, e = e.pp), 
  A = list(1, A.pp),
  effects = list(list(b0 = rep(1, nodes + nrow(murders_sp)), pop = covs$pop), 
                 list(i = 1:nodes)),
  tag = 'pp')
## fit model
pp.cov <- inla(y ~ 0 + b0 + pop + f(i, model = spde), 
  family = 'poisson', data = inla.stack.data(stk.cov), 
  control.predictor = list(A = inla.stack.A(stk.cov)), 
  E = inla.stack.data(stk.pp)$e)
## coefficients of the fixed effects
pp.cov$summary.fixed
## expected number of murders at each mesh node
ins <- which(weights != 0)
en <- exp(as.numeric(pp.cov$summary.fixed[1,1]) +
          as.numeric(pp.cov$summary.fixed[2,1])*covs$pop[1:mesh$n][ins])*weights[ins]
sum(en) ## expected number 
```


### Using `TMB` TODO

```{r using TMB}
## ensure you've run compile.stelfi()

```