# Hawkes process {#hawkes}

A  univariate Hawkes process is defined to be a self-exciting temporal point process where the conditional intensity function is given by

 $$\lambda(t) = \mu(t) + \Sigma_{i:\tau_i<t}\nu(t-\tau_i)$$ 
 where where $\mu(t)$ is the background rate of the process and $\Sigma_{i:\tau_i<t}\nu(t-\tau_i)$ is some historic temporal dependence. First introduced by @hawkes, the classic homogeneous formulation is:
 
 $$\lambda(t) = \mu + \alpha \Sigma_{i:\tau_i<t}\text{exp}(-\beta * (t-\tau_i)) $$ 
 
## The `fit_hawkes()` function

```{r}
library(stelfi)
args(fit_hawkes)
```

### Fitting a Hawkes model

A [NIWA](https://niwa.co.nz/) scientist [found a working USB in the scat of a leopard seal](https://www.nzherald.co.nz/nz/news/article.cfm?c_id=1&objectid=12201147), they then [tweeted about it](https://twitter.com/niwa_nz/status/1092610541401587712) in the hopes of finding its owner.

```{r data,message=FALSE}
data(retweets_niwa)
head(retweets_niwa)
```

```{r sort data}
## numeric time stamps
times <- unique(sort(as.numeric(difftime(retweets_niwa ,min(retweets_niwa),units = "mins"))))
```


```{r plot hist, echo = FALSE, fig.cap = "Observed counts of retweet times", fig.height = 5,fig.width = 8}
# NIWA seal first tweet according to Twitter
start <- lubridate::ymd_hms("2019-02-05 02:27:07")
## NIWA seal found owner tweet according to Twitter
end <- lubridate::ymd_hms("2019-02-07 06:50:08")
## hist
hist(retweets_niwa, breaks = "hours", axes = FALSE, 
     xlab = "", ylab = "",main = "", col = "grey",border = "grey",freq = TRUE)
axis(2,at = c(0,250))
mtext(2, line = 0, text = "Number of retweets per hour",cex = 0.8)
mtext(1,line = 1, at = c(start,end),text = c("NIWA \n tweeted","USB owner \n found"),cex = 0.9)
```



```{r fit model,results = 'hide',cache = TRUE}
params <- c(mu = 9, alpha = 3, beta = 10)
fit <- fit_hawkes(times = times, parameters = params) 
```


```{r params}
## print out estimated parameters
pars <- get_coefs(fit)
pars
```



```{r plot, echo = TRUE,fig.height = 7,fig.width = 9}
show_hawkes(fit)
```


```{r gof, echo = TRUE,fig.height = 7,fig.width = 9}
show_hawkes_GOF(fit)
```

### Fitting an ETAS-type marked model

Here we fit a univariate marked Hawkes process  where the conditional intensity function is given by

 $$\lambda(t; m(t)) = \mu + \alpha \Sigma_{i:\tau_i<t}m(\tau_i)\text{exp}(-\beta * (t-\tau_i)) $$ 
where $\mu$ is the background rate of the process and $m(t)$ is the temporal mark. Each event $i$ has an associated mark $\tau_i$ that multiples the self-exciting component of $\lambda$. 

In this example, the events are earthquakes and the marks are the Richter magnitude of each earthquake. 

```{r}
data("nz_earthquakes")
head(nz_earthquakes)
nz_earthquakes <- nz_earthquakes[order(nz_earthquakes$origintime),]
nz_earthquakes <- nz_earthquakes[!duplicated(nz_earthquakes$origintime),]
times <- nz_earthquakes$origintime
times <- as.numeric(difftime(times , min(times), units = "mins"))
marks <- nz_earthquakes$magnitude
params <- c(mu = 3, alpha = 0.05, beta = 1)
fit <- fit_hawkes(times = times, parameters = params, marks = marks) 
## print out estimated parameters
pars <- get_coefs(fit)
pars
```

```{r, echo = TRUE,fig.height = 7,fig.width = 9}
show_hawkes(fit)
```


```{r, echo = TRUE,fig.height = 7,fig.width = 9}
show_hawkes_GOF(fit)
```


## The `fit_mhawkes()` function

```{r}
args(fit_mhawkes)
```

A multivariate Hawkes process allows for between- and within-stream self-excitement. In `stelfi` the conditional intensity for the $j^{th}$ ($j = 1, ..., N$) stream is given by

 $$\lambda(t)^{j*} = \mu_j + \Sigma_{k = 1}^N\Sigma_{i:\tau_i<t} \alpha_{jk} e^{(-\beta_j * (t-\tau_i))},$$ 
where $j, k  \in (1, ..., N)$. Here, $\alpha_{jk}$ is the excitement caused by the $k^{th}$ stream on the $j^{th}$. Therefore, $\boldsymbol{\alpha}$ is an $N \times N$ matrix where the diagonals represent the within-stream excitement and the off-diagonals represent the excitement between streams.
 
```{r}
data(multi_hawkes)
```


```{r}
fit <- stelfi::fit_mhawkes(times = multi_hawkes$times, stream = multi_hawkes$stream,
                           parameters = list(mu =  c(0.2,0.2),
                                        alpha =  matrix(c(0.5,0.1,0.1,0.5),byrow = TRUE,nrow = 2),
                                        beta = c(0.7,0.7)))
get_coefs(fit)
```

## The `fit_hawkes_cbf()` function

```{r}
args(fit_hawkes_cbf)
```


### Fitting an inhomogenous Hawkes process

Here we fit a univariate *inhomogenous* marked Hawkes process where the conditional intensity function is given by

 $$\lambda(t) = \mu(t) + \alpha \Sigma_{i:\tau_i<t}\text{exp}(-\beta * (t-\tau_i)) $$ 
The background $\mu(t)$ is time varying, rather than being constant. 

The following example uses simulated data. 

```{r}
set.seed(1)
library(hawkesbow)
# Simulate a Hawkes process with mu = 1+sin(t), alpha=1, beta =2
times <- hawkesbow::hawkes(1000, fun=function(y) {1+0.5*sin(y)}, M=1.5, repr=0.5, family="exp", rate=2)$p
```

We will attempt to recover these parameter values, modelling the background as $ \mu(t) = A + Bsin(t)$. The background will be written as a function of $x$ and $y$, where $A = e^x$ and $B= logit(y) e^x$. This formulation ensures the background is never negative. 

```{r}
## The background function must take a single parameter and the time(s) at which it is evaluated
background <- function(params,times){
        A = exp(params[[1]])
        B = stats::plogis(params[[2]]) * A
        return(A + B*sin(times))
}

## The background_integral function must take a single parameter and the time at which it is evaluated
background_integral <- function(params,x){
        A = exp(params[[1]])
        B = stats::plogis(params[[2]]) * A
        return((A*x)-B*cos(x))
}
param = list(alpha = 0.5, beta = 1.5)
background_param = list(1,1)
fit <- fit_hawkes_cbf(times = times, parameters = param, background = background, background_integral = background_integral, background_parameters = background_param)
```

The estimated values of $A$ and $B$ respectively are 

```{r}
exp(fit$background_parameters[1])
plogis(fit$background_parameters[2]) * exp(fit$background_parameters[1])
```

The estimated values of $\alpha$ and $\beta$ respectively are:

```{r}
ab <- get_coefs(fit)[1:2,1]
ab
```


## The `sim_hawkes()` function

```{r}
args(sim_hawkes)
```

**`method = 1`**

```{r}
sim <- sim_hawkes(mu = 2, alpha = 0.2, beta = 0.3, plot = TRUE)
head(sim)
```

**`method = 2`**

```{r}
sim <- sim_hawkes(mu = 2, alpha = 0.2, beta = 0.3, plot = TRUE, method = 2)
head(sim)
```
