# Spatiotemporal self-exciting model {#stelfi}

Although the Hawkes process is traditionally formulated as a temporal point process, it is also possible to formulate a spatiotemporal version of the Hawkes process.

For the spatiotemporal Hawkes processes fitted by this package, temporal self-excitement follows an exponential decay function. The self-excitement over space follows a Gaussian distribution centered at the triggering event. There are two formulations of this model. The default is that the Gaussian function has a fixed covariance matrix, independent of time. Alternatively, covariance can be directly proportional to time, meaning that the self-excitement radiates out from the center over time. This can be appropriate when the mechanism causing self-excitement travels at a finite speed, but is very memory-intensive. 

The intensity function used by stelfi is:

$$\lambda(s,t) = \mu + \alpha \Sigma_{i:\tau_i<t}(\text{exp}(-\beta * (t-\tau_i)) G_i(s-x_i, t - \tau_i)) $$ 
where $\mu$ is the background rate, $\beta$ is the rate of temporal decay, $\alpha$ is the increase in intensity after an event, $\tau_i$ are the event times, $x_i$ are the event locations (in 2D Euclidean space) and $G_i(s-x_i, t - \tau_i)$ is the spatial self-excitement kernel.

For the version with time-independent spatial excitement:
$G_i(s-x_i, t - \tau_i) = N(0, \Sigma)(s - x_i)$

For the version with time-dependent spatial excitement:
$G_i(s-x_i, t - \tau_i) = N(0, (t-\tau_i)\Sigma)(s - x_i)$

## The `fit_stelfi()` function

```{r}
args(fit_stelfi)
```

## Hawkes spatial

```{r}
data(xyt, package = "stelfi")
N <- 100
locs <- data.frame(x = xyt$x[1:N], y = xyt$y[1:N])
times <- xyt$t[1:N]
domain <- sf::st_as_sf(xyt$window)
bnd <- INLA::inla.mesh.segment(as.matrix(sf::st_coordinates(domain)[, 1:2]))
smesh <- INLA::inla.mesh.2d(boundary = bnd, 
                            max.edge = 0.75, cutoff = 0.3)
param <- list(mu = 1, alpha = 3, beta = 6, xsigma = 0.2,
             ysigma = 0.2, rho = 0)
fit <- fit_stelfi(times = times, locs = locs, sf = domain, smesh = smesh, parameters = param)
get_coefs(fit)
```

## Hawkes spde

```{r}
## issues with time....
param <- list(mu = 1, alpha = 3, beta = 6, kappa = 2, tau = 1, xsigma = 0.2, ysigma = 0.2, rho = 0)
fit <- fit_stelfi(times = times, locs = locs, sf = domain, smesh = smesh, parameters = param, GMRF = TRUE)
get_coefs(fit)
res <- TMB::sdreport(fit)
show_field(res$par.random, smesh = smesh, sf = domain)
```