# Spatial log-Gaussian Cox process {#sec-lgcp}

A log-Gaussian Cox Process (LGCP) is a doubly stochastic spatial point process. In the simplest case, the intensity of the point process over space is given by
$$\Lambda(s) = \text{exp}(\beta_0 + G(s) + \epsilon)$$
where $\beta_0$ is a constant, known as the intercept, $G(s)$ is a Gaussian Markov Random Field (GMRF) and $\epsilon$ an error term.

Plotted below is a realisation of a LGCP within a disc shaped region overlain on the latent GMRF.

```{r}
#| echo: false
#| message: false
require(ggplot2)
require(stelfi)
sf <- sf::st_as_sf(spatstat.geom::disc())
smesh <- fmesher::fm_mesh_2d(loc.domain = sf::st_coordinates(sf)[, 1:2], max.edge = 0.2, cutoff = 0.1)
parameters <- c(beta = 3, log_tau = 1.6, log_kappa = 1.95)
set.seed(1)
sim <- sim_lgcp(parameters = parameters, sf = sf, smesh = smesh)
pts <- smesh$loc[rep(1:nrow(smesh$loc), times = sim$y),1:2 ]
set.seed(23)
pts <- data.frame(x = pts[,1] + runif(nrow(pts), -0.05, 0.05), y = pts[,2] + runif(nrow(pts), -0.05, 0.05))
idx <- spatstat.geom::inside.owin(pts, w = spatstat.geom::disc())
pts <- pts[idx, ]

show_field(sim$x, smesh, sf = sf, clip = TRUE) + theme_void() + theme(legend.position = "none") +
    theme(text = element_text(size=15)) +
    geom_point(data = data.frame(x = pts[,1], y = pts[,2]), aes(x = x, y = y), size = 3)

```

Following the Stochastic Partial Differential Equation (SPDE) approach proposed by @spde a Matérn covariance function is assumed for $G(s)$: 

$$C(r; \sigma, \kappa) = \frac{\sigma^2}{2^{\nu-1} \Gamma(\nu)} (\kappa \:r)^\nu K_\nu(\kappa \: r)$$


where $\kappa$ is a scaling parameter and $\sigma^2$ is the marginal variance ($\nu$ is typically fixed, see @spde). Rather than the scaling parameter, a range parameter $r=\frac{\sqrt{8}}{\kappa}$ is typically interpreted (corresponding to correlations near 0.1 at the distance $r$). The standard deviation is given as  $\sigma=\frac{1}{\sqrt{4\pi\kappa^2\tau^2}}$. In the figure above $\beta = 3$, $\text{log}(\tau) = 1.6$, and $\text{log}(\kappa) = 1.95$.

## Delauney triangluations when fitting LGCP models {#sec-mesh}

TODO

## Fitting a LGCP model

Using `spatstat` to simulate a LGCP

```{r}
#| message: false
#| warning: false
win <- spatstat.geom::disc()
## here mu is beta0
set.seed(1234)
sim <- spatstat.random::rLGCP("matern", mu = 3,
               var = 0.5, scale = 0.7, nu = 1,
               win = win)
```


```{r}
#| message: false
#| warning: false
## plotting the simulated (log) random field using ggplot
xs = attr(sim, "Lambda")$xcol; ys = attr(sim, "Lambda")$yrow
pxl <- expand.grid(x = xs, y = ys)
pxl$rf <- as.vector(attr(sim, "Lambda")$v) |> log() - 3 ## minus the intercept
require(ggplot2)
ggplot() + 
  geom_tile(data = pxl, aes(x = x, y = y, fill = rf), )  +
  labs(fill = "") + xlab("") + ylab("") + 
  scale_fill_viridis_c(option = "D", na.value = NA) +
  coord_equal() + theme_void() +
  geom_sf(data = sf::st_as_sf(win), fill = NA, linewidth = 2) +
  geom_point(data = data.frame(x = sim$x, y = sim$y), aes(x = x, y = y))
        
```

Fitting a model using `stelfi`

```{r}
#| message: false
## dataframe of point locations 
locs <- data.frame(x = sim$x, y = sim$y)
## Delauney triangulation of domain
smesh <- fmesher::fm_mesh_2d(loc = locs[, 1:2], max.edge = 0.2, cutoff = 0.1)
## sf of domain
sf <- sf::st_as_sf(win)
fit <- fit_lgcp(locs = locs, sf = sf, smesh = smesh,
                parameters = c(beta = 3, log_tau = log(1),
                               log_kappa = log(1)))
get_coefs(fit)			       
```

The estimated GMRF can be plotted using the `show_field()` function once the values have been extracted using `get_fields()`.

```{r}
#| message: false
get_fields(fit, smesh) |>
    show_field(smesh = smesh, sf = sf, clip = TRUE) + theme_void() +
    geom_point(data = data.frame(x = sim$x, y = sim$y), aes(x = x, y = y))
```
As a comparison, `inlabru` (@inlabru) is used to fit the same model to these data.


```{r}
#| message: false
require(inlabru)
require(sp)
locs_sp <- locs; sp::coordinates(locs_sp) <- c("x", "y")
domain <- as(sf, "Spatial")
matern <- INLA::inla.spde2.pcmatern(smesh, prior.sigma = c(0.7, 0.01), prior.range = c(4, 0.5) )
## latent field
cmp <- coordinates ~ random_field(coordinates, model = matern) + Intercept(1)
sp::proj4string(locs_sp) <- smesh$crs <- sp::proj4string(domain)
## fit model
fit_inla <- lgcp(cmp, locs_sp, samplers = domain, domain = list(coordinates = smesh))
pars <- rbind(fit_inla$summary.fixed[,1:2], fit_inla$summary.hyperpar[,1:2])
pars
```

Now using `spatstat`

```{r}
ss <- spatstat.model::lgcp.estK(sim, covmodel = list(model = "matern", nu = 1))
ss$clustpar
ss$modelpar
```

The table below gives the estimated parameter values from `stelfi` and `inlabru` along with the standard errors in brackets (where possible).

```{r}
#| echo: false
stelfi <- get_coefs(fit)[c(1, 4, 5),]
params <- data.frame(stelfi = c(paste(round(stelfi[,1], 3), "(", round(stelfi[,2], 3), ")")),
                     inlabru = c(paste(round(pars[,1], 3), "(", round(pars[,2], 3), ")")),
                     spatstat = c(round(ss$modelpar[3], 3), round(sqrt(8)*ss$clustpar[2], 3), round(sqrt(ss$clustpar[1]), 3)))
params <- t(params)
knitr::kable(params, col.names = c("$\\beta_0$", "$r$", "$\\sigma$"), escape = FALSE)
```

### An applied example

Using the applied example given in @stelfi a LGCP model is fitted to sasquatch sightings using the function `fit_lgcp()`. For more details on the use of the Delauney triangulation see @sec-mesh.

```{r}
#| message: false
data("sasquatch", package = "stelfi")
## get sf of the contiguous US
sf::sf_use_s2(FALSE)
us <- maps::map("usa", fill = TRUE, plot = FALSE) |>
    sf::st_as_sf() |>
    sf::st_make_valid()
## dataframe of sighting locations (lat, long)
locs <- sf::st_coordinates(sasquatch) |>
   as.data.frame()
names(locs) <- c("x", "y")
## Delauney triangulation of domain
smesh <- fmesher::fm_mesh_2d(loc = locs[, 1:2], max.edge = 2, cutoff = 1)
## fit model with user-chosen parameter starting values
fit <- fit_lgcp(locs = locs, sf = us, smesh = smesh,
                parameters = c(beta = 0, log_tau = log(1),
                               log_kappa = log(1)))
get_coefs(fit)			       
```

Again, the estimated GMRF can be plotted using the `show_field()` function once the values have been extracted using `get_fields()`.

```{r}
#| message: false
get_fields(fit, smesh) |>
    show_field(smesh = smesh, sf = us, clip = TRUE) + ggplot2::theme_classic()
```

The estimated intensity surface can be plotted using the `show_lambda()` function.

```{r}
#| message: false
show_lambda(fit, smesh = smesh, sf = us, clip = TRUE) + ggplot2::theme_classic()
```


Again, as a comparison, `inlabru` (@inlabru) is used to fit the same model to these data.


```{r}
#| message: false
require(inlabru)
require(sp)
locs_sp <- locs; sp::coordinates(locs_sp) <- c("x", "y")
domain <- as(us, "Spatial")
matern <- INLA::inla.spde2.pcmatern(smesh,
prior.sigma = c(0.1, 0.01),
prior.range = c(5, 0.01)
)
## latent field
cmp <- coordinates ~ random_field(coordinates, model = matern) + Intercept(1)
sp::proj4string(locs_sp) <- smesh$crs <- sp::proj4string(domain)
## fit model
fit_inla <- lgcp(cmp, locs_sp, samplers = domain, domain = list(coordinates = smesh))
pars <- rbind(fit_inla$summary.fixed[,1:2], fit_inla$summary.hyperpar[,1:2])
pars
```

The table below gives the estimated parameter values from `stelfi` and `inlabru` along with the standard errors in brackets.

```{r}
#| echo: false
stelfi <- get_coefs(fit)[c(1, 4, 5),]
params <- data.frame(stelfi = c(paste(round(stelfi[,1], 3), "(", round(stelfi[,2], 3), ")")),
                     inlabru = c(paste(round(pars[,1], 3), "(", round(pars[,2], 3), ")")))
params <- t(params)
knitr::kable(params, col.names = c("$\\beta_0$", "$r$", "$\\sigma$"), escape = FALSE)
```

## SPDE as GAM

TODO
