# Spatiotemporal LGCP

The LGCP model can also be used for spatiotemporal modelling where there is autoregressive temporal dependence. 

To achieve this, we choose an arbitrary number of time knots. The equation for an AR(1) process is as follows:
$$\Lambda_i(s) = \text{exp}(\beta_0 + G_i(s) + \epsilon)$$
where $i$ indexes the time knot. $\Lambda_i(s)$ is the field intensity at time knot $i$, and $G_i(s)$ the GMRF at the same time knot. Each $G_i(s)$ shares common values for $\tau$ and $\kappa$. 

Successive random fields are correlated through the formula
$$G_i(s)=\rho G_{i-1}(s) + \epsilon_i$$
where $\rho$ is a constant between -1 and +1, and $\epsilon_i$ is normally distributed with mean 0. 

```{r}
ndays <- 2
locs <- data.frame(x = xyt$x, y = xyt$y, t = xyt$t)
bnd <- INLA::inla.mesh.segment(as.matrix(sf::st_coordinates(domain)[, 1:2]))
w0 <- 2
smesh <- INLA::inla.mesh.2d(boundary = bnd,
                                max.edge = 0.75, cutoff = 0.3)
tmesh <- INLA::inla.mesh.1d(seq(0, ndays, by = w0))
fit <- fit_lgcp(locs = locs, sf = domain, smesh = smesh, tmesh = tmesh,
                    parameters = c(beta = 0, log_tau = log(1),
                                   log_kappa = log(1), atanh_rho = 0.2))
get_coefs(fit)
```


```{r}

show_lambda(fit, smesh = smesh, sf = domain, tmesh = tmesh, timestamp = 1) +
    ggplot2::theme_void() + ggplot2::ggtitle("t = 1")
show_lambda(fit, smesh = smesh, sf = domain, tmesh = tmesh, timestamp = 2) +
    ggplot2::theme_void() + ggplot2::ggtitle("t = 2")

```

## Simulating a spatiotemporal LGCP

**Option 1** using `sim_lgcp()`

```{r}
#| eval: false

parameters <- c(beta = 1, log_tau = log(1), log_kappa = log(1), atanh_rho = 0.2)
simdata <- sim_lgcp(parameters = parameters, sf = domain, smesh = smesh, tmesh = tmesh)
simdata$x[,1] |>
    show_field(smesh = smesh) +
    ggplot2::theme_void()
```

**Option 2** directly from the fitted model

```{r}
#| eval: false

simdata <- fit$simulate()
simdata$x[,1] |>
    show_field(smesh = smesh) +
    ggplot2::theme_void()
```

