# Univariate Hawkes {#sec-hawkes}

A  univariate Hawkes process (@hawkes) is defined to be a self-exciting temporal point process where the conditional intensity function is given by

 $$\lambda(t) = \mu(t) + \Sigma_{i:\tau_i<t}\nu(t-\tau_i).$$ 

Here $\mu(t)$ is the background rate of the process and $\Sigma_{i:\tau_i<t}\nu(t-\tau_i)$ is some historic temporal dependence (i.e., for times $\tau_i < t$, $ i = 1, ..., T$). The classic homogeneous formulation uses an exponential decay kernal:
 
 $$\lambda(t) = \mu + \alpha \Sigma_{i:\tau_i<t}\text{exp}(-\beta * (t-\tau_i)).$${#eq-hawkes}


Here the parameter $\alpha$ is the increase in intensity immediately after the occurrence of an event, and $\beta > 0$ controls the exponential decay of the intensity if no event has occurred. To avoid the conditional intensity heading off to infinity $\beta > \alpha$.

Plotted below is the conditional intensity ($\lambda(t), t \in [0,T]$) of a Hawkes process with $\mu = 0.1$, $\alpha = 1$, and $\beta = 1.5$. Observed events (n = 13) are shown by the vertical dashes along the x-axis. The intensity increases immediately after an event occurs and decays exponentially over time if no event is observed for some period.


```{r, fig.height = 5,fig.width = 8}
#| echo: false
library(stelfi)
library(ggplot2)
mu <- 0.5
alpha <- 1
beta <- 1.5
tim <- stelfi::sim_hawkes(mu = mu, alpha = alpha, beta = beta, n = 15)
n <- 1000
p <- seq(min(tim), max(tim), length.out = n)
lamda <- stelfi:::hawkes_intensity(mu = mu, alpha = alpha, beta = beta, times = tim, p = p)
data <-  data.frame(x = p, y = lamda)
ggplot(data = data,aes(x = x, y = y)) +
       xlab("") + ylab(expression(lambda(t))) + 
       geom_line() +  theme_minimal() + 
  geom_point(data = data.frame(x = tim, y = 0.2), pch = "|", size = 5) +
    scale_x_continuous(labels = c("t = 0","t = T"),
                       breaks = seq(min(p), max(p), length.out = 2)) 
```

## Simulating from a univariate Hawkes model {#sec-simulate-hawkes}

The function `sim_hawkes()` offers two simulation methods for simulating a realisation of a univariate Hawkes process. The default, with `method = 1`, uses the same method as algorithm 2 in @simhawkes, simulating until a given time horizon (set by argument `n`).

```{r}
#| warning: false
sim <- sim_hawkes(mu = 0.5, alpha = 1, beta = 1.5, n = 40, plot = TRUE)
```

The second option, setting `method = 2`, uses an accept/reject framework and the argument `n` specifies the number of points to simulate (default `n = 100`). 

```{r}
#| warning: false
sim <- sim_hawkes(mu = 0.5, alpha = 1, beta = 1.5, plot = TRUE, method = 2)
```


## Fitting a univariate Hawkes model

```{r}
#| eval: false
require(stelfi)
```

To fit a univariate Hawkes model in `stelfi` use the function `fit_hawkes()` with the following required arguments

   - `times` - a vector of numeric occurrence times, and 
   - `parameters` - a vector of named starting values  for $\mu$ (`mu`), $\alpha$ (`alpha`), and  $\beta$ (`beta`).

The function `get_coefs()` can then be called on the fitted model object to return the estimated parameter values.


### A simulated example

Simulating a realisation of a univariate Hawkes process (see @sec-simulate-hawkes for more details) with  $\mu = 1.3$, $\alpha = 0.4$, and $\beta = 1.5$ (over $t \in [0, 500]$).

```{r}
times <- sim_hawkes(mu = 1.3, alpha = 0.4, beta = 1.5, n = 500)
```


```{r}
#| warning: false
#| message: false
## starting values
sv <- c(mu = 1.3, alpha = 0.4, beta = 1.5)
## using stelfi
fit <- fit_hawkes(times = times, parameters = sv)
stelfi <- get_coefs(fit)
stelfi
```

As a comparison, below `emhawkes` (@emhawkes), and `hawkesbow` (@hawkesbow) are used to fit a univariate Hawkes process to the same simulated data.

```{r}
#| warning: false
#| message: false
## benchmark using emhawkes
require(emhawkes)
h <- new("hspec", mu = sv[1], alpha = sv[2], beta = sv[3])
## emhawkes requires the inter arrival times to fit the model
inter <- diff(times)
fit_em <- hfit(object = h, inter_arrival = inter)
em <- summary(fit_em)$estimate
em
## bench mark using hawkesbow
require(hawkesbow)
fit_bow <- mle(events = times, kern = "Exponential", end = max(times))
## use the Hessian to obtain the standard errors from hawkesbow
bow <- cbind(Estimate = fit_bow$par,
         "Std. error" = -fit_bow$model$ddloglik(times, max(times)) |> solve() |> diag() |> sqrt())
bow
```

The table below gives the estimated parameter values from each of `stelfi`, `emhawkes`, and `hawkesbow` along with the standard errors in brackets. Note that `hawkesbow` estimates $\frac{\alpha}{\beta}$ rather that $\beta$ directly, and that the standard errors are computed here using the returned Hessian matrix $H$ (i.e., $\sqrt{\text{diag}{-(H^{-1})}}$).

```{r}
#| echo: false
params <- data.frame(TRUTH = c(1.3, 0.4, 1.5, 0.267),
                     stelfi = c(paste(round(stelfi[,1], 3), "(", round(stelfi[,2], 3), ")"), "-"),
                     emhawkes = c(paste(round(em[,1], 3), "(", round(em[,2], 3), ")"), "-"),
                     hawkesbow = c(paste(round(bow[1,1], 3), "(", round(bow[1,2], 3), ")"), "-",
                                   paste(round(bow[3:2,1], 3), "(", round(bow[3:2,2], 3), ")")))
params <- t(params)
knitr::kable(params, col.names = c("$\\mu$", "$\\alpha$", "$\\beta$", "$\\frac{\\alpha}{\\beta}$"), escape = FALSE)
```


### An applied example

A [NIWA](https://niwa.co.nz/) scientist [found a working USB in the scat of a leopard seal](https://www.nzherald.co.nz/nz/news/article.cfm?c_id=1&objectid=12201147), they then [tweeted about it](https://twitter.com/niwa_nz/status/1092610541401587712) in the hopes of finding its owner^[This was at the time it was called Twitter (not X) so 'tweet' is the correct term!]. The dates and times of these tweets and retweets are available in `stelfi` as `retweets_niwa`.

```{r data,message=FALSE}
data(retweets_niwa)
```

The dates/times need to be numeric and sorted in ascending order (starting a time $t = 0$). Note too that there can be no simultaneous events.

```{r sort data}
## numeric time stamps
times <- unique(sort(as.numeric(difftime(retweets_niwa ,min(retweets_niwa),units = "mins"))))
```

The histogram below shows the observed counts (`r length(times)`) of unique retweet times from the original tweet ($t=0$) to the time (just over two days later) that the owner of the USB came forward, t = `r round(max(times),1)` mins.

```{r plot hist, echo = FALSE, fig.height = 5,fig.width = 8}
# NIWA seal first tweet according to Twitter
start <- lubridate::ymd_hms("2019-02-05 02:27:07")
## NIWA seal found owner tweet according to Twitter
end <- lubridate::ymd_hms("2019-02-07 06:50:08")
## hist
hist(retweets_niwa, breaks = "hours", axes = FALSE, 
     xlab = "", ylab = "",main = "", col = "grey",border = "grey",freq = TRUE)
axis(2,at = c(0,250))
mtext(2, line = 0, text = "Number of retweets per hour",cex = 0.8)
mtext(1,line = 1, at = c(start,end),text = c("NIWA \n tweeted","USB owner \n found"),cex = 0.9)
```

To fit the model chose some starting values for the parameters and supply these along with the numeric times to the function `fit_hawkes()`.

```{r fit model,results = 'hide'}
sv <- c(mu = 9, alpha = 3, beta = 10)
fit <- fit_hawkes(times = times, parameters = sv) 
```


```{r params}
## print out estimated parameters
pars <- get_coefs(fit)
pars
```
From the estimated coefficients above

 - the expected (estimated) background rate of sightings (i.e., independent sightings) is $\hat{\mu}T =$ `r round(pars[1,1],3)` $\times$ `r round(max(times),2)` $=$ `r round(pars[1,1] * max(times),2)`, which indicates that ~`r round(pars[1,1] * max(times),0)` retweets (from the `r length(times)`) were principal retweets and that the remaining were due to self-excitement;
 - the expected number of retweets “triggered” by any one retweet is estimated as $\frac{\hat{\alpha}}{\hat{\beta}}$ = `r round(pars[2,1]/pars[3,1],2)` (note the maximum this can possibly be is 1);
 - the expected number of descendants per retweet is estimated as $\frac{\hat{\beta}}{\hat{\beta} - \hat{\alpha}}$ = `r round(pars[3,1]/(pars[3,1] - pars[2,1]),2)`;
 - the rate of decay for the self-excitement is estimated as $\frac{1}{\hat{\beta}}$ = `r round(1/pars[3,1],2)`, indicating that after ~`r round(1/pars[3,1],0)` minutes a retweet is likely unrelated^[Note that unrelated is not quite the correct term here, rather that the event is likely a baseline event rather than due to self-excitement] to the previous retweet.


The `show_hawkes()` function can be called on the fitted model object to plot the estimated conditional intensity and the data, top and bottom panels below respectively.



```{r plot, echo = TRUE,fig.height = 7,fig.width = 9}
#| warning: false
#| message: false
show_hawkes(fit)
```


## Goodness-of-fit for a univariate Hawkes process {#sec-gof-hawkes}

The compensator, $\Lambda(\tilde{t})$, of any inhomogeneous Poisson process gives the expected number of events in some defined interval $[0, \tilde{t}]$ is given by

$$\Lambda(\tilde{t}) = \int_0^{\tilde{t}} \lambda(t) dt.$$

The random change theorem (@daley) states that if a set of events $[\tau_1, ..., \tau_n]$ is a realisation from a inhomogeneous Poisson process then $[\Lambda(\tau_1), ..., \Lambda(\tau_n)]$ is a realisation of a homogeneous Poisson process with unit rate. Letting $\delta \Lambda_{i} = \Lambda(\tau_i) - \Lambda(\tau_{i-1})$ for $i = 2, ..., N$ and $\delta \Lambda_{1} = \Lambda(\tau_1)$, under the theorem above $\delta \Lambda_{i} \sim \text{Exp}(1)$. Using this result a typical goodness-of-fit test is a Kolmogorov-Smirnov (KS) test (see @daley for more details) where the KS statistic is a measure of the distance between the empirical distribution of all $\delta \Lambda_{i}$ and the CDF of a $\text{Exp}(1)$ distribution.

The compensator differences can be extracted from a fitted model using the `compensator_differences()` function, and a KS test can be carried out manually:

```{r}
times <- sim_hawkes(mu = 1.3, alpha = 0.4, beta = 1.5, n = 500)
sv <- c(mu = 2, alpha = 1, beta = 5)
fit <- fit_hawkes(times = times, parameters = sv)
compensator <- compensator_differences(fit)
stats::ks.test(compensator, "pexp")
```

This gives no evidence against the compensator values coming from a $\text{Exp}(1)$ distribution.

Another goodness-of-fit test is the Box-Ljung (or Ljung–Box) test, which tests for autocorrelation between the consecutive compensator values (i.e., independence/stationarity).


```{r}
stats::Box.test(compensator, type = "Ljung")
```

This gives no evidence against the consecutive compensator values being independently distributed.

Alternatively, both tests and some diagnostic plots are returned by calling the `show_hawkes_GOF()` function. The four panels an be interpreted as follows

 - **top left**, plots the compensator values against the observed times, which under a well fitting model should align;
 - **top right**, a transformed QQ plot, the observed quantities should align with the theoretical quantiles under a well fitting model;
 - **bottom left**, the compensator differences, which under the model are assumed to be $\text{Exp}(1)$ distributed;
 - **bottom right**, consecutive compensator differences, which should show no obvious pattern (no autocorrelation evident) under a well fitting model.


```{r gof, echo = TRUE,fig.height = 7,fig.width = 9}
show_hawkes_GOF(fit)
```


## Fitting an inhomogenous Hawkes model

A univariate inhomogenous Hawkes process has $\mu = \mu(t)$ in @eq-hawkes (i.e., the baseline is time varying, rather than being constant).

Below we simulate data from a Hawkes process with $\mu(t) = [A + B*\text{sin}(\frac{2 \pi t}{365.25})]$ (with $A = 1$ and $B = 0.5$) to represent a yearly cycle and self-exciting parameters $\alpha = 1$ and $\beta = 2$  using `hawkesbow` (@hawkesbow). 

```{r}
mut <- function(t) {
 1 + 0.5*sin((2*pi*t)/365.25)
}
```

```{r}
set.seed(1)
times <- hawkesbow::hawkes(1000, fun = mut,
 M = 1.5, repr = 0.5, family = "exp", rate = 2)$p
```

To fit this model in `stelfi` the function $\mu(t)$ (`background`) and its integral $\int_0^t \mu(y) dy$ (`background_integral`) are supplied by the user. To ensure $\mu(t) > 0$ below $\mu(t)$ is written as a function of $x$ and $y$, where $A = e^x$ and $B = \text{logit}(y) e^x$.  

```{r}
background <- function(params,t){
        A = exp(params[[1]])
        B = stats::plogis(params[[2]]) * A
        return(A + B*sin((2*pi*t)/365.25))
}

background_integral <- function(params,x){
        A = exp(params[[1]])
        B = stats::plogis(params[[2]]) * A
        return((A*x)-B*cos((2*pi*x)/365.25))
}
```

These functions are then passed to the function `fit_hawkes_cbf()` alongside the observed times (`times`) and a list of starting values for both $\mu(t)$ and the self-exciting components via the argument `background_parameters` and `parameters` respectively.

```{r}
sv = list(alpha = 0.5, beta = 1.5)
background_sv = list(1,1)
fit <- fit_hawkes_cbf(times = times, parameters = sv,
background = background, background_integral = background_integral, background_parameters = background_sv)
```

The estimated values of the the transformed $A$ and $B$ parameters are returned via `fit$background_parameters`, the code below transforms them back to the original scale.

```{r}
exp(fit$background_parameters[1])
plogis(fit$background_parameters[2]) * exp(fit$background_parameters[1])
```

The estimated values of the self-excitement parameters $\alpha$ and $\beta$ are returned by the `get_coefs()` function as usual. The table below compares the estimated values to the true ones

```{r}
#| echo: FALSE
ab <- get_coefs(fit)[1:2,1]
params <- data.frame(TRUTH = c(1, 0.5, 1, 2),
                     stelfi = c(exp(fit$background_parameters[1]),
		     plogis(fit$background_parameters[2]) * exp(fit$background_parameters[1]),
		     ab[1], ab[2])
		     )
params <- t(params)
knitr::kable(params, col.names = c("$A$", "$B$", "$\\alpha$", "$\\beta$"), escape = FALSE, digit = 3)
```

The function `show_hawkes()` can be used to show the fitted model.

```{r}
#| warning: false
show_hawkes(fit)
```

The function `show_hawkes_GOF()` will show/print the model diagnostics discussed in @sec-gof-hawkes.

```{r}
#| warning: false
show_hawkes_GOF(fit)
```